\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-13 article.tex
stuff about testing framework

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
ret     & epilogue, retq & \{\}\\
br label \%l & jmp \%l & \{\}\\
br i1 \%c, label \%t, label \%f & cmpx \%c & \{\}\\
 & jz \%f & \\
 & jmp \%t & \\
 \hline
\end{tabular}
\caption{LLVM-{}- terminators to x86 translation table}\label{tab:trans-terms}
\end{table}

@@

Validating the correctness of each instruction translation is a complicated task given how low-level it is. One approach is attaching a debugger and examining the state before and after the translated sequence of x86 instructions is executed. The result of applying a binary operation,  for instance, should only affect the assigned register.

\subsection{Test Suite}

In addition to my own cases and other programs from the internet, some cases from the LLVM test suite are also used.

The structure of the test suite repository has a directory of so-called 'single source' C programs totalling to 2113. Of these only 1222 compiled successfully with \texttt{clang -S -emit-llvm} and of these only 624 parsed sucessfuly with the strip utility, of which 425 compiled successfully.

>>>>

\noindent Terminators are translated with branching instructions instead. One significant factor that needs to be accounted for is the phi nodes.
\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
ret     &  retq & \{\}\\
br label \%l & jmp \%l & \{\}\\
br i1 \%c, label \%t, label \%f & cmpx \%c & \{\}\\
 & jz \%f & \\
 & jmp \%t & \\
 \hline
\end{tabular}
\caption{LLVM-{}- terminators to x86 translation table}\label{tab:trans-terms}
\end{table}

@@

Validating the correctness of each instruction translation is a complicated task given how low-level it is. One approach is attaching a debugger and examining the state before and after the translated sequence of x86 instructions is executed. The result of applying a binary operation,  for instance, should only affect the assigned register. Another is an extensive test suite, although achieving 100\% coverage is unrealistic.

\subsection{Test Suite and  Framework}

The test suite is quite extensive although not anywhere close to complete coverage.
Some tests have been handwritten during development, mostly meant to assert one specific feature (unit tests), but others are also inteded to assert them working together (integration tests). Some of them are hand written (simple ones), others are based on the output of \texttt{clang -S -emit-llvm} stripped with the the strip utility in \texttt{bin/strip.ml}, and lastly some are generated with Python scripts to parameterize certain properties (e.g. long move-chains, many operations). In addition to this some are copied from other projects
In addition to my own cases and other programs from the internet (attributed in the file header), some cases from the LLVM test suite are also used. The structure of the test suite repository has a directory of so-called 'single source' C programs totalling to 2113. Of these only 1222 compiled successfully with \texttt{clang -S -emit-llvm} and of these only 624 parsed sucessfuly with the strip utility, of which 425 compiled successfully. In addition to these there are also the test suite used throughout Tiger development, most of which were provided by the compilers lecturers, as well as some LLVM snippets from the Dolphin compiler.

The testing framework is roughly based on the build system that came with the Tiger work environment. It is based on a collection of files that can be compiled and executed on their own. Each of these are enumerated in the test runner (\texttt{bin/testrunner.ml}) with parameters associated.   Each of these are then compiled with the \texttt{clang} compiler and executed, capturing all the relevant output data, which is stdout and exit code. After this, the same test file is compiled for each of the allocators. The executables generated are then also executed, also storing the output data.   All test cases are deterministic and will halt at some point. This is necessary because tests can easily end in an infinite loop because of erroneous translation. As each of these halt or time out, their output is compared to what was output from executables generated by the 'correct' compiler of \texttt{clang}.
